Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-09 write-ups/PLDI/paper.tex
Progress.

\newcommand{\offerfork}{\mathsf{offer\_fork}}
\newcommand{\offerkill}{\mathsf{offer\_kill}}

@@

\section{The ckill and cfork primitives}

@@

\subsection{Semantics of the new primitives}

\ADComment{General semantics: let $N$ be the maximum number of
  workgroups the GPU of interest can support [gloss over or deal with
    issue that this is kernel-dependent].  A kernel is launched with
  $M$ workgroups, for some $0 < M \leq N$.  Initially these are all
  scheduled, and are guaranteed to make independent forward progress.}

\ADComment{The purpose of ckill is for a workgroup to volunteer to
  quit execution of the kernel if the resources that it is using
  (e.g.\ the associated compute unit) are required for other tasks.  A
  ckill statement must be reached under workgroup-uniform control
  flow.  When a workgroup reaches ckill, the workgroup either
  terminates execution (because the environment requires the
  workgroup's resources), or continues execution as if the ckill
  statement were a no=op).  From a semantics perspective this is a
  non-deterministic choice: an application that invokes ckill should
  be prepared for either outcome.  In practice, the effect of ckill
  depends on environmental conditions.}

\ADComment{The purpose of cfork is for a workgroup to indicate that it
  would be useful to have additional workgroups join the computation,
  and to specify that the program point associated with the cfork
  statement would be a suitable place for a new workgroup to start
  executing.  A cfork statement must also be reached under
  workgroup-uniform control flow.  If $C$ workgroups are executing
  when a workgroup reaches a cfork statement, with $0 < C \leq M$,
  then between 0 and $M-C$ new workgroups commence execution from the
  following statement.  From a semantics perspective, the number of
  workgroups that join the computation is nondeterministic, within the
  permitted range.  In practice, how many workgroups join depends on
  environmental conditions.}

\ADComment{TODO: describe persistent variables.}

\ADComment{TODO: explain that ckill and cfork atomically update the
  kernel context (ids).}

\ADComment{TODO: somewhere we need to explain the constraints this
  places on programming: how ids work, what one can assume about them,
  etc.}

\ADComment{Now describe the more refined semantics where IDs are
  relinquished in decreasing order; explain that from a formal
  perspective thie refines the original semantics.}

\ADComment{In the above, explain at some point how a scheduler can
  exploit cfork and ckill to utilise the GPU resources.}

>>>>

\newcommand{\offerfork}{\mathsf{offer\_fork}}
\newcommand{\offerkill}{\mathsf{offer\_kill}}
\newcommand{\globalbarrier}{\mathsf{global\_barrier}}
\newcommand{\resizingglobalbarrier}{\mathsf{resizing\_global\_barrier}}
\newcommand{\getgroupid}[1]{\mathsf{get\_group\_id(#1)}}
\newcommand{\getnumgroups}[1]{\mathsf{get\_num\_groups(#1)}}

@@

\section{The Cooperative Kernels Programming Model}

We describe our cooperative kernels programming model as an extension
to the industry-standard OpenCL programming model for heterogenous
many-core architectures.  However, the cooperative kernels concept is
more general, and could be applied to extend other GPU programming
models, such as CUDA and HSA.  \ADComment{However, unclear that we could implement it on top of these, but that's not really the point.}

\subsection{Semantics of Cooperative Kernels}

As with a regular OpenCL kernel, a cooperative kernel is launched by
the host application, with a specified number of workgroups, each
consisting of a specified number of threads.  The kernel's behaviour
is described by a kernel function, written in OpenCL C but using the
additional language extensions that we present below, and a number of
parameters to the kernel function are passed by the host application.

The parameters to a cooperative kernel are \emph{immutable}, as if
they were qualified with the \texttt{const} keyword (though pointer
parameters can refer to mutable data).

A regular OpenCL kernel can be executed by a multi-dimensional grid of
workgroups, called an ND-range.  In our current proposal, the ND-range
for a cooperative kernel must be one-dimensional.  \ADComment{Say why?
  And mention that none of the applications that we deem would benefit
  from cooperative kernels need a multi-dimensional ND-range.}

As in regular OpenCL, each workgroup can be multi-dimensional.

\paragraph{Restored variables}

A variable declared in the root scope of the cooperative kernel can
optionally be annotated with a new \texttt{restore} qualifier.  By
annotating a variable with \texttt{restore}, the programmer indicates
that when new workgroups are spawned via $\offerkill$, the threads in
these workgroups should obtain a well-defined initial value for the
variable, obtained from the workgroup that executed $\offerkill$.  We
detail the semantics for this when we describe $\offerkill$ below.

Only private memory variables can be annotated with \texttt{restore},
which may include pointers to local or global memory so long as the
pointer variables themselves are stored in private memory.  In
particular, variables and arrays in local memory cannot be annotated
with \texttt{restore}.  \ADComment{The preceding is what we have
  implemented, but actually it is pretty clear how we could handle
  local memory restoration if we wanted to; it's just that none of our
  examples needed it.}

\paragraph{Active workroups}

If the host application launches a cooperative kernel specifying that
$N$ workgroups should execute the kernel, then when the kernel
commences execution it should be executed by some number of workgroups
$M$, with $1\leq M \leq N$.  The requested number of workgroups is
called the \emph{workgroup limit} for the cooperative kernel, and $M$
is the number of \emph{active workgroups}.  As explained below, the
number $M$ of active workgroups may change during the lifetime of a
cooperative kernel.

Fair scheduling is guaranteed between active workgroups.  That is, if
some thread in an active workgroup is enabled then eventually some
thread in the active workgroup is guaranteed to execute an
instruction.  Note that our programming model does not mandate that
threads within a workgroup must be fairly scheduled, and indeed this
is often not the case in GPU programming models, because threads are
often organised into \emph{warps} or \emph{wavefronts} within a
workgroup, exhibiting lock-step predicated execution that is
inherently unfair.

The initial workgroups have consecutive workgroup ids in the range
$[0, M-1]$.

The OpenCL primitive $\getgroupid{0}$ retrieves the id of a
workgroup as usual.  When executed by a cooperative kernel, the OpenCL
primitive $\getnumgroups{0}$ returns $M$, the current number
of active workgroups.  Because the ND-ranges of cooperative kernels
are one-dimensional, $\getgroupid{\mathit{i}}$ and
$\getnumgroups{\mathit{i}}$ return 0 and 1, respectively, for $i >
0$.

\ADComment{What happens to $M$ when a workgroup reaches the end of a
  cooperative kernel?  Feels like $M$ should decrease, but what does
  this mean for barrier-only kernels?  Do they need to account for the
  number of active workgroups changing during the final barrier
  interval?  An alternative is to say that there is an implicit global barrier at kernel exit.}

\paragraph{Semantics for $\offerkill$}

Our first additional language primitive, $\offerkill$, is a
workgroup-level function: similar to the OpenCL intra-workgroup
$\mathsf{barrier}$ primitive, if $\offerkill$ appears inside
conditional code then it must be executed by \emph{all} threads in a
workgroup, with all threads having executed the same number of
iterations of any enclosing loops.

Suppose a workgroup with id $m$ executes $\offerkill$.  If $m < M-1$,
i.e.\ the workgroup's id is not the largest among the active
workgroups, then $\offerkill$ is a no-op.  If instead $m = M-1$, there
is a non-deterministic choice between (a) $\offerkill$ executing as a
no-op, or (b) execution of the workgroup ceasing and the number of
active workgroups $M$ being decremented by one.  Case (b) corresponds
to the scheduler accepting the workgroup's offer to cease execution,
allowing the scheduler to schedule some other task on the compute unit
on which the workgroup was executing.

The semantics of $\offerkill$ is atomic.  That is, if multiple
workgroups reach $\offerkill$ simultaneously, the semantics is as if
each workgroup executes $\offerkill$ in some arbitrary sequential
order.  \ADComment{Should we say something about linearisability?}

\paragraph{Semantics for $\offerfork$}

Our second new language primitive, $\offerfork$, is also a workgroup-level
function.

Suppose a workgroup with id $m\leq M$ executes $\offerfork$.  An
integer $k$ in the range $[0, N-M]$ is chosen nondeterministically,
$k$ new workgroups are spawned with consecutive ids in the range
$[M, M+k-1]$, and the active workgroup count $M$ is incremented by
$k$.

The $k$ new workgroups commence execution at the program point
immediately following the $\offerfork$ call.  The private variables of
each thread in a new workgroup are all uninitialised, as are any local
variables and arrays shared across the workgroup; reading from these
locations without initialising them is an undefined behaviour.  There
are two exceptions to this:

\begin{itemize}

\item because the parameters to a cooperative kernel are immutable,
  the new threads have access to these parameters and can safely read
  from them;

\item for each variable $v$ annoated with \texttt{restore}, every new
  thread's copy of $v$ is initialised to the value that thread 0 in
  workgroup $m$ held for $v$ at the point of the $\offerfork$ call.

\end{itemize}

The second exception, for restored variables, can be stated more
formally as follows.  Let $\mathit{WG}[x][y].v$ denote the value
stored in variable $v$ of thread $y$ in workgroup $x$.  Then for each
new workgroup $i \in [M, M+k-1]$ and each thread $j$ in workgroup $i$,
$\mathit{WG}[i][j].v$ is initialised to $\mathit{WG}[m][0].v$ for each
variable $v$ annotated with \texttt{restore}.

Notice that $k=0$ is always a legitimate choice for the number of
workgroups to be spawned by an $\offerfork$ call, and if the number of
active workgrouups $M$ is equal to the workgroup limit $N$, $k=0$ is
guaranteed.

\ADComment{Remark on our decision to have every thread inherit the
  state of the master thread from workgroup $m$.}

\paragraph{Global barriers}

Many irregular algorithms require synchronisation across workgroups.
This can be achieved via a \emph{global barrier}, but a global barrier
is not provided as a primitive in OpenCL, CUDA or HSA.  This is
because, as discussed in Section~\ref{ }, an inter-workgroup barrier
requires fair scheduling of workgroups.  As a result, past works have
either split computation into multiple separate kernels, so that the
end of one kernel and the start of the next kernel implicitly provides
a global barrier~\cite{ }, or have assumed the occupancy-bound
execution model to build a global barrier that works reliably on
today's GPU platforms.

Because workgroups of a cooperative kernel are fairly scheduled, a
global barrier primitive can be provided.  We specify two such primitives: $\globalbarrier()$
and $\resizingglobalbarrier()$.

A $\globalbarrier()$ is a kernel-level function: if it appears in
conditional code then it must be reached by \emph{all} threads
executing the cooperative kernel.  On reaching a $\globalbarrier()$, a
thread waits until all threads have arrived at the barrier.  Once all
threads have arrived, the threads may proceed past the barrier with
the guarantee that all global memory accesses issued before the
barrier have completed.

A $\resizingglobalbarrier()$ is identical to a global barrier, except
that, semantically, every workgroup executes $\offerkill$ on reaching
the barrier, and every workgroup executes $\offerfork$ on leaving the
barrier.  Thus $\resizingglobalbarrier()$ is semantically equivalent to:
%
\lstset{basicstyle=\tt}
\begin{lstlisting}
  $\offerkill()$;
  $\globalbarrier()$;
  $\offerfork()$;
\end{lstlisting}
\lstset{basicstyle=\scriptsize\tt}

The $\globalbarrier()$ primitive can be implemented directly using the
atomic operations provided by the OpenCL 2.0 memory model~\cite{...},
but the implementation is involved, hence why we include this function
as a primitive.  As noted above, the $\resizingglobalbarrier()$
primitive is, semantically, synactic sugar for a sequence of other
primitive calls.  We include it as a primitive because (a) we have
found that global barriers often provide a natural point for
cooperation with the scheduler, and (b) identifying a known point at
which workgroups are required to synchronize and at which workgroups
can be safely killed and forked enables a custom implementation of
$\resizingglobalbarrier()$ that provides the semantics described
above, but is significantly more efficient than the above sequence of
calls.  We discuss our efficient implementation of
$\resizingglobalbarrier$ in detail in Section~\ref{...}.

\subsection{Backwards compatibility}

\ADComment{Perhaps this would fit better under ``programming guidelines'' -- not sure.}

\ADComment{Make the point here that if an irregular algorithm is
  written as a cooperative kernel then it will execute just as safely
  as it does today on current hardware if: (1) the \texttt{restore}
  annotation and $\offerkill$/$\offerfork$ calls are defined away to
  nothing, (2) $\resizingglobalbarrier$ is redefined to
  $\globalbarrier$, (3) a standard implementation of $\globalbarrier$
  that assumes the occupancy-bound execution model is provided.  If the kernel uses blocking features---the global barrier, mutexes, etc., then it may not execute safely due to the possibility of unfair scheduling, but it will execute no less safely than it would if the cooperative kernels style were not used.}

\subsection{Programming guidelines}

An OpenCL programmer writing a cooperative kernel should pay careful
attention to the following details.

\paragraph{Keeping at least one workgroup alive}

The programer is responsible for keeping at least one workgroup alive.  In particular, if all workgroups execute $\offerkill$, the scheduler is at liberty to destroy all workgroups, causing the cooperative kernel to exit.  This can be guarded against by wrapping each $\offerkill$ call in a conditional statement:

\lstset{basicstyle=\tt}
\begin{lstlisting}
  if($\getgroupid{0}$ != 0) $\offerkill$;
\end{lstlisting}
\lstset{basicstyle=\scriptsize\tt}

We could, alternatively, have formulated the semantics of $\offerkill$
so that workgroup 0 is never killed.  However, there might be a use
case for a cooperative kernel reaching a point where it would be
acceptable for the kernel to exit, although desirable for some
remaining computation to be performed if competing workloads allow it.

\paragraph{A changing number of workgroups}

In regular OpenCL, $\getnumgroups{0}$ returns a runtime-constant
value: once the number of executing workgroups has been identified, it
is guaranteed not to change.

In contrast, the number of workgroups executing a cooperative kernel
can vary due to workgroups executing $\offerkill$, $\offerfork$ and
$\resizingglobalbarrier$.  The programmer must therefore write their
cooperative kernel in a manner that is robust to changes in the value
returned by $\getnumgroups{0}$.

The main impact of this relates to partitioning of data to be
processed.

\ADComment{Tyler, can we give an example of a data processing loop
  that would have to be changed?}

\paragraph{Cooperative kernels that only use barriers}

\ADComment{Need to account for the issue that $M$ might decrease as
  workgroups actually exit the kernel.}

If a cooperative kernel does not directly issue $\offerkill$ and
$\offerfork$ calls then the only possibilty for the number of
workgroups to change is due to $\resizingglobalbarrier$ being invoked.

In this case, because the resizing barrier is a kernel-level function,
the statements executed by the cooperative kernel can be divided into
\emph{resizing barrier intervals}: a resizing barrier interval starts
at a resizing barrier or the kernel entry point, and ends at a
resizing barrier or a kernel exit point.  At any moment of execution,
all threads are guaranteed to be executing within the same resizing
barrier interval.  As a result, the number of workgroups executing the
kernel is guaratneed to remain unchanged within each resizing barrier
interval, so the result returned by $\getnumgroups{0}$ is guaratneed
to be identical on repeated calls during such an interval.  This
guarantee may be exploited for efficiently when writing a cooperative
kernel that uses global barriers but does not call $\offerfork$ and
$\offerkill$ directly.

\subsection{Non-functional Requirements of Cooperative Kernels}

The semantics presented in Section~\ref{...} describe the envelope of
behaviour that a developer of a cooperative kernel should be prepared
for.

However, the aim of cooperative kernels is to find a balance that
allows \emph{efficient execution} of long-running kernels
(e.g.\ implementing irregular data-parallel algorithms), and
\emph{responsive multi-tasking} so that the GPU can be shared between
one or more cooperative kernels and a number of short, high-priority
tasks with soft real-time constraints.

To achieve this balance, both an implementation of the ooperative
kernels model, and the programmer of a cooperative kernel, must strive
to meet a number of non-functional requirements, which we outline
here.

\paragraph{Sufficient $\offerkill$ calls}

Recall that the purpose of $\offerkill$ is to provide the scheduler
with an opportunity to destroy a workgroup in order to schedule
higher-priority tasks.  Because these higher-priority tasks, such as
graphics rendering, may have soft real-time constraints, the scheduler
relies on the cooperative kernel to execute $\offerkill$ sufficiently
frequently that the scheduler can meet such real-time constraints.

\ADComment{Got here.}

@@
